---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}

# knitr ops
knitr::opts_chunk$set(
   eval=FALSE, 
   include=TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# load lib
library(pineconer)

```

# Pineconer

<!-- badges: start -->
<!-- badges: end -->

The goal of pineconer is to provide a way to store and access vectors using the Pincone API

## Installation

You can install the development version of pineconer like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

## Configuration

You can set the API key and region in your env file. usethis::edit_r_environ() is a convenient way to open the file. Next set the following two variables.

PINECONE_API_KEY = "YOUR API KEY"
PINECONE_ENVIRONMENT = "PINECONE Region"


## Collection Operations

Pincone API support the following index operations, the are implemented like for like in the pineconer package. 

```{r index}

# list collections
  list_collections()

# create collection
  create_collection("xxx","xxx")
  
# describe collection
  describe_collection(collection_name = "xxx")
  
# delete collection
  delete_collection("xxx")

```

## Index Operations

```{r}
# list indexes
  list_indexes()

# create index
  create_index("xxx")

# describe index
  describe_index("xxx")

# delete index
  delete_index("xxx")
  
# configure index
 configure_index(index_name = "xxx", replica  = 1, pod_type = "s1.x2")
  
  
```

