---
title: "Getting Started with pineconer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pineconer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction
The `pineconer` package provides an R interface to the [Pinecone Vector Database](https://www.pinecone.io/), a managed vector database designed for machine learning applications. This vignette demonstrates how to use `pineconer` for common vector database operations using the classic iris dataset.

## Setup

### Installation

```{r install}
# Install from GitHub (when available)
# remotes::install_github("bob-rietveld/pineconer")
```

### API Key Configuration

Before using `pineconer`, you need to set your Pinecone API key. Add it to your `~/.Renviron` file:

```
PINECONE_API_KEY=your_api_key_here
```

Then restart R or run:

```{r renviron}
readRenviron("~/.Renviron")
```

Load the package:

```{r setup}
library(pineconer)
```

## Working with Indexes

### List Existing Indexes

```{r list-indexes}
# List all indexes in your project
indexes <- list_indexes()
print(indexes$content)
```

### Create an Index

Pinecone supports two types of indexes: serverless and pod-based.

#### Serverless Index (Recommended)

```{r create-serverless}
# Create a serverless index for storing 4-dimensional vectors (like iris features)
result <- create_index(
  name = "iris-demo",
  dimension = 4,
  metric = "cosine",
  spec = list(
    serverless = list(
      cloud = "aws",
      region = "us-east-1"
    )
  )
)
print(result$status_code)  # 201 on success
```

#### Pod-Based Index

```{r create-pod}
# Alternative: Create a pod-based index
result <- create_index(
  name = "iris-pod-demo",
  dimension = 4,
  metric = "cosine",
  spec = list(
    pod = list(
      environment = "us-east-1-aws",
      pod_type = "p1.x1",
      pods = 1
    )
  )
)
```

### Describe an Index

```{r describe-index}
# Get index details including the host for data operations
index_info <- describe_index("iris-demo")
print(index_info$content$host)
print(index_info$content$status)
```

## Vector Operations with the Iris Dataset

The iris dataset contains 150 observations with 4 numeric features, making it ideal for demonstrating vector operations.

### Preparing Data

```{r prepare-data}
# Load iris dataset
data(iris)

# Prepare vectors for upserting
vectors <- lapply(1:nrow(iris), function(i) {
  list(
    id = paste0("iris-", i),
    values = as.numeric(iris[i, 1:4]),
    metadata = list(
      species = as.character(iris$Species[i]),
      sepal_length = iris$Sepal.Length[i],
      sepal_width = iris$Sepal.Width[i]
    )
  )
})

# Preview first vector
print(vectors[[1]])
```

### Upsert Vectors

```{r upsert}
# Upsert vectors in batches (Pinecone recommends batches of 100)
batch_size <- 100
n_batches <- ceiling(length(vectors) / batch_size)

for (i in 1:n_batches) {
  start_idx <- (i - 1) * batch_size + 1
  end_idx <- min(i * batch_size, length(vectors))

  result <- vector_upsert(
    index = "iris-demo",
    vectors = vectors[start_idx:end_idx]
  )

  cat("Batch", i, "- Status:", result$status_code, "\n")
}
```

### Query Vectors

Find similar flowers based on their measurements:
```{r query}
# Query using a sample vector (first iris observation)
query_vector <- as.numeric(iris[1, 1:4])

results <- vector_query(
  index = "iris-demo",
  vector = query_vector,
  top_k = 5,
  include_metadata = TRUE
)

# Results are returned as a tidy tibble by default
print(results$content)
```

### Query with Metadata Filters

```{r query-filter}
# Find similar vectors but only among setosa species
results <- vector_query(
  index = "iris-demo",
  vector = query_vector,
  top_k = 5,
  filter = list(species = list(`$eq` = "setosa"))
)

print(results$content)
```

### Fetch Specific Vectors

```{r fetch}
# Fetch vectors by ID
fetched <- vector_fetch(
  index = "iris-demo",
  ids = c("iris-1", "iris-2", "iris-3")
)

print(fetched$content)
```

### Update Vector Metadata

```{r update}
# Update metadata for a specific vector
result <- vector_update(
  index = "iris-demo",
  vector_id = "iris-1",
  meta_data = list(
    species = "setosa",
    sepal_length = 5.1,
    verified = TRUE
  )
)

print(result$status_code)  # 200 on success
```

### Get Index Statistics

```{r stats}
# Get statistics about the index
stats <- describe_index_stats("iris-demo")

print(stats$content$totalVectorCount)
print(stats$content$dimension)
```

### Delete Vectors

```{r delete}
# Delete specific vectors
result <- vector_delete(
  index = "iris-demo",
  ids = c("iris-1", "iris-2")
)

# Delete all vectors in a namespace
result <- vector_delete(
  index = "iris-demo",
  delete_all = TRUE,
  name_space = "test-namespace"
)
```

## Working with Collections

Collections are static snapshots of an index that can be used to create new indexes.

### Create a Collection

```{r create-collection}
# Create a collection from an existing index
result <- create_collection(
  name = "iris-backup",
  source = "iris-demo"
)

print(result$status_code)
```

### List Collections

```{r list-collections}
collections <- list_collections()
print(collections$content)
```

### Describe a Collection

```{r describe-collection}
collection_info <- describe_collection("iris-backup")
print(collection_info$content)
```

## Working with Namespaces

Namespaces allow you to partition vectors within an index:

```{r namespaces}
# Upsert to a specific namespace
result <- vector_upsert(
  index = "iris-demo",
  vectors = vectors[1:50],
  name_space = "training"
)

result <- vector_upsert(
  index = "iris-demo",
  vectors = vectors[51:100],
  name_space = "validation"
)

# Query a specific namespace
results <- vector_query(
  index = "iris-demo",
  vector = query_vector,
  top_k = 5,
  name_space = "training"
)
```

## Cleanup

```{r cleanup}
# Delete the index when done
delete_index("iris-demo")

# Delete the collection
delete_collection("iris-backup")
```

## Error Handling

All functions return a consistent response structure:

```{r error-handling}
result <- describe_index("non-existent-index")

if (result$status_code != 200) {
  cat("Error:", result$status_code, "\n")
  # Access raw response for details
  print(httr::content(result$http))
} else {
  print(result$content)
}
```

## Tips and Best Practices

1. **Batch Operations**: When upserting many vectors, use batches of 100 for optimal performance.

2. **Dimension Consistency**: Ensure all vectors have the same dimension as defined when creating the index.

3. **Metadata**: Use metadata for filtering queries. Common patterns include categories, timestamps, and source identifiers.

4. **Namespaces**: Use namespaces to logically separate data (e.g., by user, environment, or data source).

5. **Index Hosting**: Store the index host from `describe_index()` to avoid repeated API calls.

## Additional Resources

- [Pinecone Documentation](https://docs.pinecone.io/)
- [Pinecone API Reference](https://docs.pinecone.io/reference/api/introduction)
- [Vector Embeddings Guide](https://www.pinecone.io/learn/vector-embeddings/)
